Reversible Regime Change: Climate-driven phytoplankton community shifts in the Cariaco Basin, Venezuela

### 2025-08-04 agujournaltemplate.tex
Ago final comments

\authors{Benjamin Post\affil{1,2}, Esteban Acevedo-Trejos\affil{3}, Subhendu Chakraborty\affil{1}, Andrew D. Barton\affil{4}, Agostino Merico\affil{1}}

@@

\affiliation{1}{Systems Ecology Group, Leibniz Centre for Tropical Marine Research (ZMT), Bremen, Germany}
\affiliation{2}{School of Science, Constructor University, Bremen, Germany}
\affiliation{3}{Earth Surface Process Modelling, GFZ German Research Centre for Geosciences, Potsdam, Germany}
\affiliation{4}{Scripps Institution of Oceanography and Department of Ecology, Behavior and Evolution, University of California San Diego, La Jolla, CA, United States}
%(repeat as many times as is necessary)

@@

\begin{keypoints}
\item The CARIACO phytoplankton community time series reveals two clusters, separated by a shift from high- to low-upwelling conditions after 2004 and a return to high-upwelling conditions following 2013.
\item No significant differences are observed in diversity metrics between the two clusters, with the exception of a lower turnover and a higher community evenness during low-upwelling conditions.
\item Gradient forest analysis identifies the AMO climate index as the strongest predictor of community change, highlighting the influence of large-scale climatic oscillations on this tropical coastal ecosystem.
\end{keypoints}

@@

Phytoplankton communities are integral to oceanic biogeochemical cycles and are sensitive indicators of climate-driven environmental variability.
Long-term time series can capture this variability, allowing us to untangle the effects of environmental change on local communities. This study investigates changes in the phytoplankton community of the Cariaco Basin, Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data. To investigate impacts of larger scale climate, we considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both of which have been linked to variations in wind that drive upwelling in the Cariaco Basin.
Cluster analysis identified two distinct community states associated with a shift from high to low upwelling conditions in 2004, followed by a return to a community and conditions similar to those before the shift in 2014. In contrast, community diversity declined between 1998 and 2009, independent of changes in the upwelling regime.
To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that the AMO index was the strongest predictor of change, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2. Our study emphasizes the impact of long-term climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the resilient nature of the tropical marine ecosystem under global change.

@@

% Topic Overview
Phytoplankton are microscopic algae that form the foundation of the ocean food web. Because they respond quickly to environmental changes, studying them can show how marine ecosystems are affected by climate.
% Paper Overview
This study looked at how phytoplankton communities in the Cariaco Basin, off the coast of Venezuela, changed from 1995 to 2017. We compared microscopy cell counts and monthly water samples with weather data and two major climate indicators: the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both linked to local weather and currents.
% Findings Summary
The results showed a major shift in the phytoplankton community in 2004, linked to a decrease in upwelling, with conditions partially returning to the previous state by 2014. Interestingly, community diversity declined between 1998 and 2009, although upwelling had not yet changed. Further analysis showed that the AMO index, along with sea surface temperature, MEI v.2 and nitrate levels in the water column, were the strongest predictors of changes in the phytoplankton community.
% Key Takeaways
These findings contribute to a broader understanding of how climate variability shapes marine phytoplankton communities and underscore the importance of sustained long-term observations in detecting and interpreting ecological responses to climate change.

@@

% Phytoplankton is important
Phytoplankton provide the foundation for marine food webs and play crucial roles in biogeochemical processes \cite{falkowski_biogeochemical_1998}. Their short generation times and high assemblage diversity allow phytoplankton communities to respond relatively quickly to environmental changes, making them suitable indicators of ecosystem state and dynamics \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}. Long-term time series of biological observations are crucial for understanding how phytoplankton communities respond to both short- and long-term changes in the physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}. Such time series are rare, particularly in the tropics \cite{clarke_does_2017}.

@@

% PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
However, the shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensities from 2004 to 2013 did not appear to be a permanent change in regime. Data from 2013 to 2017, the final years of the CARIACO (CArbon Retention In A Colored Ocean) time series, indicated a return to strong upwelling intensities and high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}. The impact of large-scale climatic processes on the observed shifts in the phytoplankton community remains unclear. The Atlantic Multidecadal Oscillation (AMO) has emerged as an important factor influencing the climate system in the region and the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean Sea \cite{stephenson_changes_2014}. The El Niño Southern Oscillation (ENSO) is teleconnected to the Caribbean Sea, given that during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures increase and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. However, using the time series available only until 2010, \citeA{taylor_ecosystem_2012} found no significant correlation between MEI v.2 and in situ observations, and did not test for correlations with the AMO index.

% What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
\citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation". Previous analyses of phytoplankton community diversity in the Cariaco Basin did not include time series data beyond 2011, and no detailed investigations have yet been conducted on changes in diversity based on microscopy cell counts. Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 to 2017, 2) changes in the structure of the phytoplankton community data and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community. Our analysis focuses exclusively on bottom-up drivers of community changes because these data are available throughout the entire time series and at a monthly resolution. Our study builds on and substantially expands the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. In doing so, we hope to provide a more comprehensive assessment of long-term phytoplankton community dynamics in the Cariaco Basin, offering new insights into the ecological responses to environmental variability across decadal scales.

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin situated on the northeastern Venezuelan continental shelf, with a sill to the north running almost parallel to the coastline at a mean depth of around \qty{100}{m}. This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at the BATS and HOT sites, resulting in the formation of anoxic conditions below \qty{\sim 275}{m} depth.

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Map_CAR.pdf}
\caption{Bathymetric map of the CARIACO Basin in the Caribbean Ocean. The station is located at \ang{10.50}N, \ang{64.67}W).}
\label{fig:map}
\end{figure}

\subsection{In-situ biological and environmental data}
    The CARIACO Ocean Time Series Program collected biological and environmental data, which we used here to examine the connections between environmental and ecological change in the region. Samples were collected on monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.67}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, along with other deployments and samples not utilized in this work. Details about sampling procedures, analytical methods, data quality control, and inter-calibration procedures have already been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

@@

    All in situ data, including the Niskin bottle \cite{mullerkarger2019niskin}, CTD \cite{mullerkarger2019ctd}, and phytoplankton taxonomy \cite{troccoli2019phytoplankton}, were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}. 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (no more than six months) in coverage. In total, we used a complete set of 254 monthly data points from 1995 to 2017, for which all variables were available.

\subsection{Meteorological and climate data}
    %from Taylor et al. 2012
    In addition to the in situ data, we also examine meteorological and climatic data that are known to influence ocean conditions and marine ecosystems. Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}. We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the monthly averaged ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}. For simplicity, we refer to this variable as "wind speed" throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate of total precipitation, comprising all water that falls on Earth's surface as rain or snow.

@@

\subsection{Data analysis}    
    In the analysis, we set out to find if there is a structure in the distribution and temporal evolution of diversity metrics and the community composition, as well as to identify potential bottom-up drivers of this change. 
    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores ($z = \frac{(x – \mu)}{\sigma}$, where $x$ is the yearly mean, $\mu$ is the mean and $\sigma$ is the standard deviation of all yearly means). Data from 1995, 2016, and 2017 were not included because insufficient coverage (i.e. more than six missing monthly samples) in these years prevented us from producing reliable yearly estimates.

@@

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers in the Cariaco Basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. We used monthly genus level abundances integrated over the top \qty{100}{\meter} as indicator variables and limited the data to only those genera that were counted more than 10 times over the entire time series. This produced a data set of 193 monthly samples and 89 observed genera abundances, where all predictor variables were also available. The genus abundances were transformed by the natural logarithm plus one prior to model runs. The drivers/predictor variables considered in the analysis were the AMO and MEI v.2 indices, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO_3]}$, $\mathrm{[PO_4]}$ and $\mathrm{[SiO_4]}$). All predictor variables are in some way related to the local climate and upwelling system and could be potential bottom-up drivers of community change, some more directly (e.g. the isotherm and nutrient concentrations) and some more indirectly, such as the climate indices and meteorological data. 
    The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. Gradient forest runs were performed with 1500 bootstrapped trees and a correlation threshold of 0.5.
    We show a model run without any time lags in the supplementary material (Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we tested for an improved predictive power of time lags of up to 6 months for all climate variables, as well as 3-month lags for in situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix). All the code and data used in this analysis and to create the plots are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.

    \begin{table}
    \caption{Data and corresponding units used as predictor variables for the gradient forest analysis.}
    \centering
    \resizebox{\textwidth}{!}{\begin{tabular}[t]{rrl}
    \toprule
    Variable Name & Units  & Data source \\
    \midrule
    MEI v.2 & unitless & Climate index provided by NOAA \\
    AMO & unitless & Climate index provided by NSF NCAR \\
    Wind speed & \qty{}{\meter/\second} & ERA5 model reanalysis, negative 10 m u-component of wind speed \\
    Precipitation & \qty{}{\meter} & ERA5 model reanalysis, total precipitation\\
    Evaporation & \qty{}{\meter} & ERA5 model reanalysis, evaporation \\
    SST & \qty{}{\celsius} & CTD, Average temperature measured down to \qty{10}{\meter} depth\\
    \qty{21}{\celsius} isotherm & \qty{}{\meter} & CTD, depth at which temperature reaches 21°C\\
    Salinity & PSU & Niskin, interpolated down to \qty{100}{\meter} depth \\
    $\mathrm{NO}_3$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
    $\mathrm{PO}_4$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
    $\mathrm{SiO}_4$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
    \end{tabular}}
    \label{tab:DataSources}
    \end{table}

@@

\subsection{Interannual changes in the environment and ecosystem}
    % Figure 2
    Coordinated, interannual changes are observable in normalized yearly means of all environmental variables, as well as log-transformed chlorophyll \textit{a} and phytoplankton cell counts per functional group (Figure \ref{fig:zscore}). Distinct patterns emerge in wind speed, evaporation, precipitation, SST, and the \qty{21}{\celsius} isotherm depth. Periods of positive wind anomalies (blue) correspond to negative anomalies in SST and upwelling (red, indicated by a shallower isotherm depth), reflecting stronger upwelling conditions. These patterns are consistent with the Atlantic Multidecadal Oscillation (AMO), where a positive AMO phase is correlated with weaker upwelling regimes (see also Figure~\ref{fig:sup:correlation} in the Appendix). Yearly mean nutrient concentrations show strong positive anomalies at the beginning of the time series and are generally well correlated. Nitrate and phosphate anomalies tended to be higher during periods of stronger upwelling but remained relatively low between 2002 and 2014, with nitrate showing more interannual variability. Silicate anomalies, in contrast, did not follow the same pattern as nitrate and phosphate. Instead, elevated concentrations were observed between 2008 and 2013, potentially driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced consumption due to lower diatom abundance.
    Chlorophyll \textit{a} and functional group cell counts show positive anomalies throughout most of the first half of the time series, followed by a significant reduction after 2004, suggesting a shift. This shift occurs during the period between 2003 and 2013, when the AMO index presents mostly positive anomalies, while upwelling-related variables indicate a weak upwelling regime (low wind, high SST, deep isotherm). Although data are sparse, nutrients and biological variables show a strong increase following 2013, coinciding with a return to negative anomalies in the AMO index and more upwelling-favorable conditions.
    The annual means of diversity metrics do not follow this trend, but are reduced between 1998 and 2008. Shannon diversity specifically appears elevated past 2006 and the general pattern that emerges is that mean diversity is lower in years, where mean chlorophyll \textit{a} is high. However, in 1996 and 1997 positive anomalies in cell counts, chlorophyll \textit{a} and diversity metrics coincide.

    \begin{figure}
    \noindent\includegraphics[width=\textwidth]{fig/Figure2_ZScores_v2.pdf}
    \caption{Time series showing annual z-scores of selected environmental, climatological, and ecological variables. Red indicates negative annual anomalies and blue indicates positive annual anomalies. Deeper shades indicate larger anomalies. Chlorophyll and functional group counts were log-transformed prior to normalization. The years 1995 (n=2) and 2016 (n=5), 2017 (n=1) were removed due to insufficient monthly data to generate robust annual means. Additionally we marked years with two or three and four or more missing monthly values.}
    \label{fig:zscore}
    \end{figure}

@@

\section{Gradient Forest Analysis supplement}

>>>>

\authors{Benjamin Post\affil{1,2}, Esteban Acevedo-Trejos\affil{3}, Subhendu Chakraborty\affil{1}, Andrew D. Barton\affil{4}, Agostino Merico\affil{1,2,5}}

@@

\affiliation{1}{Systems Ecology Group, Leibniz Centre for Tropical Marine Research (ZMT), Bremen, Germany}
\affiliation{2}{School of Science, Constructor University, Bremen, Germany}
\affiliation{3}{Earth Surface Process Modelling, GFZ German Research Centre for Geosciences, Potsdam, Germany}
\affiliation{4}{Scripps Institution of Oceanography and Department of Ecology, Behavior and Evolution, University of California San Diego, La Jolla, CA, United States}
\affiliation{5}{Faculty of Biology and Chemistry, University of Bremen, Germany}
%(repeat as many times as is necessary)

@@

\begin{keypoints}
\item The CARIACO phytoplankton community time series is characterized by two distinct temporal clusters separated by a shift in 2004 from high to low upwelling and by a return to high upwelling conditions in 2014.
\item No significant differences are observed in diversity metrics between the two clusters, with the exception of a lower turnover and a higher community evenness during low upwelling conditions.
\item Gradient forest analysis identifies the AMO climate index as the strongest predictor of community change, highlighting the influence of large-scale climatic oscillations on this tropical coastal ecosystem.
\end{keypoints}

@@

Phytoplankton communities are integral to oceanic biogeochemical cycles and are sensitive indicators of climate-driven environmental variability.
Long-term time series can capture this variability, allowing us to unravel the effects of environmental change on local communities. This study investigates changes in the phytoplankton community of the Cariaco Basin, off the coast of Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data. To investigate impacts of larger-scale climate variability, we considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both of which have been linked to variations in wind that drive upwelling in the Cariaco Basin.
Cluster analysis identified two distinct community states separated by a shift from high to low upwelling conditions in 2004. In 2014, community and environmental conditions returned to a state similar to that before the shift in 2004. In contrast, community diversity declined between 1998 and 2009, independent of changes in the upwelling regime.
To quantify the effects of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that the AMO index was the strongest predictor of change, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2. Our study emphasizes the impact of long-term climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and suggests a cyclic (if not resilient) behavior of this tropical marine ecosystem under global change.

@@

% Topic Overview
Phytoplankton are photosynthesizing unicellular microorganisms that form the foundation of the ocean food web. Since they respond quickly to environmental changes, studying them can show how marine ecosystems are affected by climate.
% Paper Overview
This study investigated how phytoplankton communities in the Cariaco Basin, off the coast of Venezuela, changed from 1995 to 2017. We compared microscopy cell counts and monthly water samples with weather data and two major climate indicators: the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both linked to local weather and currents.
% Findings Summary
The results showed a major shift in the phytoplankton community in 2004, linked to a decrease in upwelling, with conditions partially returning in 2014 to the state antecedent 2004. Interestingly, community diversity declined between 1998 and 2009 independently from changes in the upwelling regime. Further analysis showed that the AMO index, along with sea surface temperature, MEI v.2 and nitrate concentrations in the water column, were the strongest predictors of changes in phytoplankton community.
% Key Takeaways
These findings contribute to a broader understanding of how climate variability shapes marine phytoplankton communities and underscore the importance of sustained long-term observations in detecting and interpreting ecological responses to climate change.

@@

% Phytoplankton is important
Phytoplankton provide the foundation for marine food webs and play crucial roles in biogeochemical processes \cite{falkowski_biogeochemical_1998}. Their short generation times and high taxonomic and functional group diversity allow phytoplankton communities to respond relatively quickly to environmental changes, making them suitable indicators of ecosystem state and dynamics \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}. Long-term time series of biological observations are crucial for understanding how phytoplankton communities respond to both short- and long-term changes in the physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}. Such time series are rare, particularly in the tropics \cite{clarke_does_2017}.

@@

% PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
The shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensities from 2004 to 2013 did not appear to be a permanent change in regime. Data from 2014 to 2017, the final years of the CARIACO (CArbon Retention In A Colored Ocean) time series, indicated a return to strong upwelling intensities and high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}. The impact of large-scale climatic processes on the observed shifts in the phytoplankton community remains unclear. The Atlantic Multidecadal Oscillation (AMO) has emerged as an important factor influencing the climate system in the region and the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean Sea \cite{stephenson_changes_2014}. The El Niño Southern Oscillation (ENSO) is teleconnected to the Caribbean Sea, given that during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures increase and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. However, using the time series available only until 2010, \citeA{taylor_ecosystem_2012} found no significant correlation between MEI v.2 and in situ observations, and did not test for correlations with the AMO index.

% What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
\citeA{taylor_ecosystem_2012} noted that the observed changes could not be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation". But, to date, all analyses of phytoplankton community diversity in the Cariaco Basin do not include time series data beyond 2011, and no detailed investigations have yet been conducted on changes in diversity based on microscopy cell counts. Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 to 2017, 2) changes in the structure of the phytoplankton community and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community. Our analysis focuses exclusively on bottom-up drivers of community changes because these data are available throughout the entire time series and at a monthly resolution. Our study builds on and substantially expands the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. In doing so, we aim to provide a more comprehensive assessment of long-term phytoplankton community dynamics in the Cariaco Basin, offering new insights into the ecological responses to environmental variability across decadal scales.

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
    %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin situated on the northeastern Venezuelan continental shelf, with a sill to the north running almost parallel to the coastline at a mean depth of around \qty{100}{m}. This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at the BATS and HOT sites, resulting in the formation of anoxic conditions below a depth of  \qty{\sim 275}{m}.

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/Map_CAR.pdf}
\caption{Bathymetric map of the CARIACO Basin in the Caribbean Ocean. The CARIACO mooring station is located at \ang{10.50}N, \ang{64.67}W).}
\label{fig:map}
\end{figure}

\subsection{In-situ biological and environmental data}
    The CARIACO Ocean Time Series Program collected biological and environmental data, which we used here to examine the connections between environmental and ecological change in the region. Samples were collected on monthly research cruises at a mooring station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.67}W) \cite{muller-karger_scientific_2019}.
    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, along with other deployments and samples not utilized in this work. Details about sampling procedures, analytical methods, data quality control, and inter-calibration procedures have already been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

@@

    All in situ data, including the Niskin bottle \cite{mullerkarger2019niskin}, CTD \cite{mullerkarger2019ctd}, and phytoplankton taxonomy \cite{troccoli2019phytoplankton}, were obtained from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}. 
    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (no more than six months) in coverage. In total, we used a complete set of 254 monthly data points from 1995 to 2017, for which all variables were available.

\subsection{Meteorological and climate data}
    %from Taylor et al. 2012
    In addition to the in situ data, we also examined meteorological and climatic data that are known to influence ocean conditions and marine ecosystems. Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}. We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the monthly averaged ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}. For simplicity, we refer to this variable as "wind speed" throughout the article. The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell. Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate of total precipitation, comprising all water that falls on Earth's surface as rain or snow.

@@

\subsection{Data analysis}    
    In the analysis, we set out to find if there was a structure in the distribution and temporal evolution of diversity metrics and community composition and, additionally, to identify potential bottom-up drivers of this change. 
    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores ($z = \frac{(x – \mu)}{\sigma}$, where $x$ is the yearly mean, $\mu$ is the mean and $\sigma$ is the standard deviation of all yearly means). Data from 1995, 2016, and 2017 were not included because insufficient coverage (i.e. more than six missing monthly samples) in these years prevented us from producing reliable yearly estimates.

@@

    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
    To quantify the effects of bottom-up drivers on shifts in the phytoplankton community of the Cariaco Basin, we employed a Gradient Forest analysis. Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}. Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor. The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover. Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}. We used monthly genus level abundances integrated over the top \qty{100}{\meter} as indicator variables and limited the data to only those genera that were counted more than 10 times over the entire time series. This produced a data set of 193 monthly samples and 89 observed genera abundances, where all predictor variables were also available. The genus abundances were transformed by the natural logarithm plus one prior to model runs. The drivers/predictor variables considered in the analysis were the AMO and MEI v.2 indices, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO_3]}$, $\mathrm{[PO_4]}$ and $\mathrm{[SiO_4]}$). All predictor variables are in some way related to the local climate and upwelling system and could be potential bottom-up drivers of community change, some more directly (e.g. the isotherm and nutrient concentrations) and some more indirectly, such as the climate indices and meteorological data. 
    The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}. Gradient Forest runs were performed with 1500 bootstrapped trees and a correlation threshold of 0.5.
    We show a model run without any time lags in the supplementary material (Figure \ref{fig:sup:GFoutput_nolags}). To consider the temporal lag in effect, particularly of climate indices, we tested for an improved predictive power of time lags of up to 6 months for all climate variables, as well as 3-month lags for in situ measurements in exploratory runs. We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix). All the codes and data used in this analysis and used to create the plots are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.

    \begin{table}
    \caption{Data and corresponding units used as predictor variables for the Gradient Forest analysis.}
    \centering
    \resizebox{\textwidth}{!}{\begin{tabular}[t]{rrl}
    \toprule
    Variable Name & Units  & Data source \\
    \midrule
    MEI v.2 & unitless & Climate index provided by NOAA \\
    AMO & unitless & Climate index provided by NSF NCAR \\
    Wind speed & \qty{}{\meter/\second} & ERA5 model reanalysis, negative 10 m u-component of wind speed \\
    Precipitation & \qty{}{\meter} & ERA5 model reanalysis, total precipitation\\
    Evaporation & \qty{}{\meter} & ERA5 model reanalysis, evaporation \\
    SST & \qty{}{\celsius} & CTD, Average temperature measured down to \qty{10}{\meter} depth\\
    \qty{21}{\celsius} isotherm & \qty{}{\meter} & CTD, depth at which temperature reaches 21°C\\
    Salinity & PSU & Niskin, interpolated down to \qty{100}{\meter} depth \\
    $\mathrm{[NO_3]}$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
    $\mathrm{[PO_4]}$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
    $\mathrm{[SiO_4]}$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
    \end{tabular}}
    \label{tab:DataSources}
    \end{table}

@@

\subsection{Interannual changes in the environment and ecosystem}
    % Figure 2
    Coordinated, interannual changes are observable in normalized yearly means of all environmental variables, in log-transformed chlorophyll \textit{a}, and in phytoplankton cell counts per functional group (Figure \ref{fig:zscore}). Distinct patterns emerge in wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth. Periods of positive wind anomalies (blue) correspond to negative anomalies in SST and upwelling (red, indicated by a shallower isotherm depth), reflecting stronger upwelling conditions. These patterns are consistent with the Atlantic Multidecadal Oscillation (AMO), where a positive AMO phase is correlated with weaker upwelling regimes (see also Figure~\ref{fig:sup:correlation} in the Appendix). Yearly mean nutrient concentrations show strong positive anomalies at the beginning of the time series and are generally well correlated. Nitrate and phosphate anomalies tended to be higher during periods of stronger upwelling but remained relatively low between 2002 and 2014, with nitrate showing more interannual variability. Silicate anomalies, in contrast, did not follow the same pattern as nitrate and phosphate. Instead, elevated concentrations were observed between 2008 and 2013, potentially driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced consumption due to lower diatom abundance.
    Chlorophyll \textit{a} and functional group cell counts show positive anomalies throughout most of the first half of the time series, followed by a significant reduction after 2004, suggesting a shift. This shift occurs during the period between 2003 and 2013, when the AMO index presents mostly positive anomalies, while upwelling-related variables indicate a weak upwelling regime (low wind, high SST, deep isotherm). Although data are sparse, nutrients and biological variables show a strong increase from 2014, coinciding with a return to negative anomalies in the AMO index and to more upwelling-favorable conditions.
    The annual means of diversity metrics do not follow this trend, but are reduced between 1998 and 2008. Shannon diversity specifically appears elevated past 2006 and the general pattern that emerges is that mean diversity is lower in years when mean chlorophyll \textit{a} is high. However, in 1996 and 1997 positive anomalies in cell counts, chlorophyll \textit{a} and diversity metrics coincide.

    \begin{figure}
    \noindent\includegraphics[width=\textwidth]{fig/Figure2_ZScores_v2.pdf}
    \caption{Time series showing annual z-scores of selected environmental, climatological, and ecological variables. Red indicates negative annual anomalies and blue indicates positive annual anomalies. Deeper shades indicate larger anomalies. Chlorophyll and functional group counts were log-transformed prior to normalization. The years 1995 (n=2) and 2016 (n=5), 2017 (n=1) were removed due to insufficient monthly data to generate robust annual means. Additionally, we marked years with two or three and four or more missing monthly values.}
    \label{fig:zscore}
    \end{figure}

@@

\section{Gradient Forest analysis supplement}
